---
title: "Hand-on exercise 5 Visual Analytics"
date: "Feb 14,2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
output:
  quarto::quarto_html_document:
    self_contained: true
editor: visual
---

# **1.  Creating Ternary Plot with R**

## **1.1 Overview**

Ternary plots are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) It’s display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

In this hands-on, you will learn how to build ternary plot programmatically using R for visualising and analysing population structure of Singapore.

The hands-on exercise consists of four steps:

-   Install and launch **tidyverse** and **ggtern** packages.

-   Derive three new measures using *mutate()* function of **dplyr** package.

-   Build a static ternary plot using *ggtern()* function of **ggtern** package.

-   Build an interactive ternary plot using *plot-ly()* function of **Plotly R** package.

## **1.2 Installing and launching R packages**

For this exercise, two main R packages will be used in this hands-on exercise, they are:

-   [**ggtern**](http://www.ggtern.com/), a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

We will also need to ensure that selected **tidyverse** family packages namely: **readr**, **dplyr** and **tidyr** are also installed and loaded.

In this exercise, version 3.2.1 of **ggplot2** will be installed instead of the latest version of **ggplot2**. This is because the current version of **ggtern** package is not compatible to the latest version of **ggplot2**.

The code chunks below will accomplish the task.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## **1.3 Data Preparation**

### **1.3.1 The data**

For the purpose of this hands-on exercise, the [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) data will be used. The data set has been downloaded and included in the data sub-folder of the hands-on exercise folder. It is called *respopagsex2000to2018_tidy.csv* and is in csv file format.

### **1.3.2 Importing Data**

To important *respopagsex2000to2018_tidy.csv* into R, ***read_csv()*** function of **readr** package will be used.

```{r}
#Reading the data into R environment
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

### **1.3.3 Preparing the Data**

Next, use the ***mutate()*** function of **dplyr** package to derive three new measures, namely: young, active, and old.

```{r}
#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

## **13.4 Plotting Ternary Diagram with R**

### **13.4.1 4.1 Plotting a static ternary diagram**

Use ***ggtern()*** function of **ggtern** package to create a simple ternary plot.

```{r}
#Building the static ternary plot
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
#Building the static ternary plot 
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) + geom_point() + labs(title="Population structure, 2015") + theme_rgbw()
```

### **1.4.2 Plotting an interative ternary diagram**

The code below create an interactive ternary plot using ***plot_ly()*** function of **Plotly R**.

```{r}
# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

# **2  Heatmap for Visualising and Analysing Multivariate Data**

## **2.1 Overview**

Heatmaps visualise data through variations in colouring. When applied to a tabular format, heatmaps are useful for cross-examining multivariate data, through placing variables in the columns and observation (or records) in rowa and colouring the cells within the table. Heatmaps are good for showing variance across multiple variables, revealing any patterns, displaying whether any variables are similar to each other, and for detecting if any correlations exist in-between them.

In this hands-on exercise, you will gain hands-on experience on using R to plot static and interactive heatmap for visualising and analysing multivariate data.

## **2.2 Installing and Launching R Packages**

Before you get started, you are required to open a new Quarto document. Keep the default html as the authoring format.

Next, you will use the code chunk below to install and launch **seriation**, **heatmaply**, **dendextend** and **tidyverse** in RStudio.

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

## **2.3 Importing and Preparing The Data Set**

In this hands-on exercise, the data of [World Happines 2018 report](https://worldhappiness.report/ed/2018/) will be used. The data set is downloaded from [here](https://s3.amazonaws.com/happiness-report/2018/WHR2018Chapter2OnlineData.xls). The original data set is in Microsoft Excel format. It has been extracted and saved in csv file called **WHData-2018.csv**.

### **2.3.1 Importing the data set**

In the code chunk below, **read_csv()** of *readr* is used to import WHData-2018.csv into R and parsed it into tibble R data frame format.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

The output tibbled data frame is called **wh**.

### **2.3.2 Preparing the data**

Next, we need to change the rows by country name instead of row number by using the code chunk below

```{r}
row.names(wh) <- wh$Country
```

Notice that the row number has been replaced into the country name.

### **2.3.3 Transforming the data frame into a matrix**

The data was loaded into a data frame, but it has to be a data matrix to make your heatmap.

The code chunk below will be used to transform *wh* data frame into a data matrix.

```{r}
wh1 <- dplyr::select(wh, c(3, 7:12))
wh_matrix <- data.matrix(wh)
```

Notice that **wh_matrix** is in R matrix format.

## **2.4 Static Heatmap**

There are many R packages and functions can be used to drawing static heatmaps, they are:

-   [heatmap()](https://www.rdocumentation.org/packages/stats/versions/3.6.0/topics/heatmap)of R stats package. It draws a simple heatmap.

-   [heatmap.2()](https://www.rdocumentation.org/packages/gplots/versions/3.0.1.1/topics/heatmap.2) of **gplots** R package. It draws an enhanced heatmap compared to the R base function.

-   [pheatmap()](https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12/topics/pheatmap) of [**pheatmap**](https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12) R package. **pheatmap** package also known as Pretty Heatmap. The package provides functions to draws pretty heatmaps and provides more control to change the appearance of heatmaps.

-   [**ComplexHeatmap**](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) package of R/Bioconductor package. The package draws, annotates and arranges complex heatmaps (very useful for genomic data analysis). The full reference guide of the package is available [here](https://jokergoo.github.io/ComplexHeatmap-reference/book/).

-   [**superheat**](https://cran.r-project.org/web/packages/superheat/) package: A Graphical Tool for Exploring Complex Datasets Using Heatmaps. A system for generating extendable and customizable heatmaps for exploring complex datasets, including big data and data with multiple data types. The full reference guide of the package is available [here](https://rlbarter.github.io/superheat/).

In this section, you will learn how to plot static heatmaps by using **heatmap()** of *R Stats* package.

### **2.4.1 heatmap() of R Stats**

In this sub-section, we will plot a heatmap by using *heatmap()* of Base Stats. The code chunk is given below.

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      Rowv=NA, Colv=NA)
```

***Note:***

-   By default, **heatmap()** plots a cluster heatmap. The arguments ***Rowv=NA*** and ***Colv=NA*** are used to switch off the option of plotting the row and column dendrograms.

To plot a cluster heatmap, we just have to use the default as shown in the code chunk below.

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

**Note:**

-   The order of both rows and columns is different compare to the native wh_matrix. This is because heatmap do a reordering using clusterisation: it calculates the distance between each pair of rows and columns and try to order them by similarity. Moreover, the corresponding dendrogram are provided beside the heatmap.

Here, red cells denotes small values, and red small ones. This heatmap is not really informative. Indeed, the Happiness Score variable have relatively higher values, what makes that the other variables with small values all look the same. Thus, we need to normalize this matrix. This is done using the *scale* argument. It can be applied to rows or to columns following your needs.

The code chunk below normalises the matrix column-wise.

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6, 
                      cexCol = 0.8,
                      margins = c(10, 4))
```

Notice that the values are scaled now. Also note that **margins** argument is used to ensure that the entire x-axis labels are displayed completely and, cexRow and cexCol arguments are used to define the font size used for y-axis and x-axis labels respectively.

## **2.5 Creating Interactive Heatmap**

[**heatmaply**](http://talgalili.github.io/heatmaply/) is an R package for building interactive cluster heatmap that can be shared online as a stand-alone HTML file. It is designed and maintained by Tal Galili.

Before we get started, you should review the [Introduction to Heatmaply](https://cran.r-project.org/web/packages/heatmaply/vignettes/heatmaply.html) to have an overall understanding of the features and functions of Heatmaply package. You are also required to have the [user manual](https://cran.r-project.org/web/packages/heatmaply/heatmaply.pdf)of the package handy with you for reference purposes.

In this section, you will gain hands-on experience on using **heatmaply** to design an interactive cluster heatmap. We will still use the wh_matrix as the input data.

### **2.5.1 Working with heatmaply**

```{r}
heatmaply(mtcars)
```

The code chunk below shows the basic syntax needed to create n interactive heatmap by using **heatmaply** package.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

Note that:

-   Different from *heatmap()*, for *heatmaply()* the default horizontal dendrogram is placed on the left hand side of the heatmap.

-   The text label of each raw, on the other hand, is placed on the right hand side of the heat map.

-   When the x-axis marker labels are too long, they will be rotated by 135 degree from the north.

### **2.5.2 Data trasformation**

When analysing multivariate data set, it is very common that the variables in the data sets includes values that reflect different types of measurement. In general, these variables’ values have their own range. In order to ensure that all the variables have comparable values, data transformation are commonly used before clustering.

Three main data transformation methods are supported by *heatmaply()*, namely: scale, normalise and percentilse.

#### 2.5.2.1 Scaling method

-   When all variables are came from or assumed to come from some normal distribution, then scaling (i.e.: subtract the mean and divide by the standard deviation) would bring them all close to the standard normal distribution.

-   In such a case, each value would reflect the distance from the mean in units of standard deviation.

-   The *scale* argument in *heatmaply()* supports column and row scaling.

The code chunk below is used to scale variable values columewise.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

#### 2.5.2.2 Normalising method

-   When variables in the data comes from possibly different (and non-normal) distributions, the normalize function can be used to bring data to the 0 to 1 scale by subtracting the minimum and dividing by the maximum of all observations.

-   This preserves the shape of each variable’s distribution while making them easily comparable on the same “scale”.

Different from Scaling, the normalise method is performed on the input data set i.e. wh_matrix as shown in the code chunk below.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

#### 2.5.2.3 Percentising method

-   This is similar to ranking the variables, but instead of keeping the rank values, divide them by the maximal rank.

-   This is done by using the ecdf of the variables on their own values, bringing each value to its empirical percentile.

-   The benefit of the percentize function is that each value has a relatively clear interpretation, it is the percent of observations that got that value or below it.

Similar to Normalize method, the Percentize method is also performed on the input data set i.e. wh_matrix as shown in the code chunk below.

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### **2.5.3 Clustering algorithm**

**heatmaply** supports a variety of hierarchical clustering algorithm. The main arguments provided are:

-   *distfun*: function used to compute the distance (dissimilarity) between both rows and columns. Defaults to dist. The options “pearson”, “spearman” and “kendall” can be used to use correlation-based clustering, which uses as.dist(1 - cor(t(x))) as the distance metric (using the specified correlation method).

-   *hclustfun*: function used to compute the hierarchical clustering when *Rowv* or *Colv* are not dendrograms. Defaults to *hclust*.

-   *dist_method* default is NULL, which results in “euclidean” to be used. It can accept alternative character strings indicating the method to be passed to distfun. By default *distfun* is “dist”” hence this can be one of “euclidean”, “maximum”, “manhattan”, “canberra”, “binary” or “minkowski”.

-   *hclust_method* default is NULL, which results in “complete” method to be used. It can accept alternative character strings indicating the method to be passed to *hclustfun*. By default hclustfun is hclust hence this can be one of “ward.D”, “ward.D2”, “single”, “complete”, “average” (= UPGMA), “mcquitty” (= WPGMA), “median” (= WPGMC) or “centroid” (= UPGMC).

In general, a clustering model can be calibrated either manually or statistically.

### **2.5.4 Manual approach**

In the code chunk below, the heatmap is plotted by using hierachical clustering algorithm with “Euclidean distance” and “ward.D” method.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

### **2.5.5 Statistical approach**

In order to determine the best clustering method and number of cluster the *dend_expend()* and *find_k()* functions of **dendextend** package will be used.

First, the *dend_expend()* will be used to determine the recommended clustering method to be used.

```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

The output table shows that “average” method should be used because it gave the high optimum value.

Next, *find_k()* is used to determine the optimal number of cluster.

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

Figure above shows that k=3 would be good.

With reference to the statistical analysis results, we can prepare the code chunk as shown below.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

### **2.5.6 Seriation**

One of the problems with hierarchical clustering is that it doesn’t actually place the rows in a definite order, it merely constrains the space of possible orderings. Take three items A, B and C. If you ignore reflections, there are three possible orderings: ABC, ACB, BAC. If clustering them gives you ((A+B)+C) as a tree, you know that C can’t end up between A and B, but it doesn’t tell you which way to flip the A+B cluster. It doesn’t tell you if the ABC ordering will lead to a clearer-looking heatmap than the BAC ordering.

**heatmaply** uses the seriation package to find an optimal ordering of rows and columns. Optimal means to optimize the Hamiltonian path length that is restricted by the dendrogram structure. This, in other words, means to rotate the branches so that the sum of distances between each adjacent leaf (label) will be minimized. This is related to a restricted version of the travelling salesman problem.

Here we meet our first seriation algorithm: Optimal Leaf Ordering (OLO). This algorithm starts with the output of an agglomerative clustering algorithm and produces a unique ordering, one that flips the various branches of the dendrogram around so as to minimize the sum of dissimilarities between adjacent leaves. Here is the result of applying Optimal Leaf Ordering to the same clustering result as the heatmap above.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

The default options is “OLO” (Optimal leaf ordering) which optimizes the above criterion (in O(n\^4)). Another option is “GW” (Gruvaeus and Wainer) which aims for the same goal but uses a potentially faster heuristic.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

The option “mean” gives the output we would get by default from heatmap functions in other packages such as gplots::heatmap.2.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

The option “none” gives us the dendrograms without any rotation that is based on the data matrix.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none")
```

### **2.5.7 Working with colour palettes**

The default colour palette uses by **heatmaply** is *viridis*. heatmaply users, however, can use other colour palettes in order to improve the aestheticness and visual friendliness of the heatmap.

In the code chunk below, the Blues colour palette of rColorBrewer is used

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors = Blues)
```

### **2.5.8 The finishing touch**

Beside providing a wide collection of arguments for meeting the statistical analysis needs, *heatmaply* also provides many plotting features to ensure cartographic quality heatmap can be produced.

In the code chunk below the following arguments are used:

-   *k_row* is used to produce 5 groups.

-   *margins* is used to change the top margin to 60 and row margin to 200.

-   *fontsizw_row* and *fontsize_col* are used to change the font size for row and column labels to 4.

-   *main* is used to write the main title of the plot.

-   *xlab* and *ylab* are used to write the x-axis and y-axis labels respectively.

    ```{r}
    heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
              Colv=NA,
              seriate = "none",
              colors = Blues,
              k_row = 5,
              margins = c(NA,200,60,NA),
              fontsize_row = 4,
              fontsize_col = 5,
              main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
              xlab = "World Happiness Indicators",
              ylab = "World Countries"
              )
    ```

# **3  Visual Multivariate Analysis with Parallel Coordinates Plot**

## **3.1 Overview**

Parallel coordinates plot is a data visualisation specially designed for visualising and analysing multivariate, numerical data. It is ideal for comparing multiple variables together and seeing the relationships between them. For example, the variables contribute to Happiness Index. Parallel coordinates was invented by [Alfred Inselberg](http://www.math.tau.ac.il/~aiisreal/) in the 1970s as a way to visualize high-dimensional data. This data visualisation technique is more often found in academic and scientific communities than in business and consumer data visualizations. As pointed out by [Stephen Few](https://www.perceptualedge.com/articles/b-eye/parallel_coordinates.pdf)(2006), “This certainly isn’t a chart that you would present to the board of directors or place on your Web site for the general public. In fact, the strength of parallel coordinates isn’t in their ability to communicate some truth in the data to others, but rather in their ability to bring meaningful multivariate patterns and comparisons to light when used interactively for analysis.” For example, parallel coordinates plot can be used to characterise clusters detected during customer segmentation.

By the end of this hands-on exercise, you will gain hands-on experience on:

-   plotting statistic parallel coordinates plots by using *ggparcoord()* of **GGally** package,

-   plotting interactive parallel coordinates plots by using **parcoords** package, and

-   plotting interactive parallel coordinates plots by using **parallelPlot** package.

## **3.2 Installing and Launching R Packages**

For this exercise, the **GGally**, **parcoords**, **parallelPlot** and **tidyverse** packages will be used.

The code chunks below are used to install and load the packages in R.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

## **3.3 Data Preparation**

In this hands-on exercise, the World Happinees 2018 (http://worldhappiness.report/ed/2018/) data will be used. The data set is download at https://s3.amazonaws.com/happiness-report/2018/WHR2018Chapter2OnlineData.xls. The original data set is in Microsoft Excel format. It has been extracted and saved in csv file called **WHData-2018.csv**.

In the code chunk below, `read_csv()` of **readr** package is used to import *WHData-2018.csv* into R and save it into a tibble data frame object called *wh*.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

## **3.4 Plotting Static Parallel Coordinates Plot**

In this section, you will learn how to plot static parallel coordinates plot by using [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of **GGally** package. Before getting started, it is a good practice to read the function description in detail.

### **3.4.1 Plotting a simple parallel coordinates**

Code chunk below shows a typical syntax used to plot a basic static parallel coordinates plot by using `ggparcoord()`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

Notice that only two argument namely `data` and `columns` is used. `Data` argument is used to map the data object (i.e. *wh*) and `columns` is used to select the columns for preparing the parallel coordinates plot.

### **3.4.2 Plotting a parallel coordinates with boxplot**

The basic parallel coordinates failed to reveal any meaning understanding of the World Happiness measures. In this section, you will learn how to makeover the plot by using a collection of arguments provided by `ggparcoord()`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

Things to learn from the code chunk above.

-   `groupColumn` argument is used to group the observations (i.e. parallel lines) by using a single variable (i.e. Region) and colour the parallel coordinates lines by region name.

-   `scale` argument is used to scale the variables in the parallel coordinate plot by using `uniminmax` method. The method univariately scale each variable so the minimum of the variable is zero and the maximum is one.

-   `alphaLines` argument is used to reduce the intensity of the line colour to 0.2. The permissible value range is between 0 to 1.

-   `boxplot` argument is used to turn on the boxplot by using logical `TRUE`. The default is `FALSE`.

-   `title` argument is used to provide the parallel coordinates plot a title.

### **3.4.3 Parallel coordinates with facet**

Since `ggparcoord()` is developed by extending **ggplot2** package, we can combination use some of the ggplot2 function when plotting a parallel coordinates plot.

In the code chunk below, `facet_wrap()` of **ggplot2** is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

One of the aesthetic defect of the current design is that some of the variable names overlap on x-axis.

### **3.4.4 Rotating x-axis text label**

To make the x-axis text label easy to read, let us rotate the labels by 30 degrees. We can rotate axis text labels using `theme()` function in ggplot2 as shown in the code chunk below

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

Thing to learn from the code chunk above:

-   To rotate x-axis text labels, we use `axis.text.x` as argument to `theme()` function. And we specify `element_text(angle = 30)` to rotate the x-axis text by an angle 30 degree.

### **3.4.5 Adjusting the rotated x-axis text label**

Rotating x-axis text labels to 30 degrees makes the label overlap with the plot and we can avoid this by adjusting the text location using `hjust` argument to theme’s text element with `element_text()`. We use `axis.text.x` as we want to change the look of x-axis text.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

## **3.5 Plotting Interactive Parallel Coordinates Plot: parallelPlot methods**

[**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) is an R package specially designed to plot a parallel coordinates plot by using ‘htmlwidgets’ package and [d3.js](https://d3js.org/). In this section, you will learn how to use functions provided in parallelPlot package to build interactive parallel coordinates plot.

### **3.5.1 The basic plot**

The code chunk below plot an interactive parallel coordinates plot by using `parallelPlot()`.

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

Notice that some of the axis labels are too long. You will learn how to overcome this problem in the next step.

### **3.5.2 Rotate axis label**

In the code chunk below, `rotateTitle` argument is used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

One of the useful interactive feature of parallelPlot is we can click on a variable of interest, for example Happiness score, the monotonous blue colour (default) will change a blues with different intensity colour scheme will be used.

### **3.5.3 Changing the colour scheme**

We can change the default blue colour scheme by using `continousCS` argument as shown in the code chunl below.

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

### **3.5.4 Parallel coordinates plot with histogram**

In the code chunk below, `histoVisibility` argument is used to plot histogram along the axis of each variables.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

## **3.6 References**

-   [*ggparcoord()*](http://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](http://ggobi.github.io/ggally/) package

-   [**parcoords** user guide](https://cran.r-project.org/web/packages/parcoords/parcoords.pdf)

-   [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/parallelPlot.pdf)
