---
title: "Take-home exercise 4 Visual Analytics"
date: "Mar 8,2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
  message: false
output:
  quarto::quarto_html_document:
    self_contained: true
editor: visual
---

# **1  Overview**

For this Take Home Exercise, there are several task needs to be done to create the Shiny application.

-   To evaluate and determine the necessary R packages needed for your Shiny application are supported in R CRAN,

-   To prepare and test the specific R codes can be run and returned the correct output as expected,

-   To determine the parameters and outputs that will be exposed on the Shiny applications, and

-   To select the appropriate Shiny UI components for exposing the parameters determine above.

This submission includes the prototype report for the group project, which will includes:

-   the data preparation process,

-   the selection of data visualisation techniques used,

-   and the data visualisation design and interactivity principles and best practices implemented.

# **2  Project Infomation**

The purpose of this project is to create a Shiny app with user-friendly interface and functions, and also create a website for user to discover the historical Weather changes in Singapore.

# **3  Get Started (Load R Packages)**

```{r}
pacman::p_load(sf,terra,gstat,tmap,viridis,tidyverse,dplyr)
```

# 4 . Data Preparation

import the merged data of all the stations across 10 years (2014-2023)

```{r}
rfdata <- read.csv('data/merged_data.csv')
```

Import the data that includes the latitude and longitude of the weather stations in Singapore.

```{r}
rfstation <- read.csv('data/RainfallStation.csv')

rfstation <- rfstation %>%
  rename(Station = Station.Code)
```

Join the latitude and longitude to the rfdata, so that in the analysis part we can map the geospatial analysis.

```{r}
rfdata<- rfdata  %>%
  left_join(rfstation, by = "Station")
```

# **5  Exploratory Data Analysis**

# **6  Clustering**
